---
title: "Explore Dataset & "
author: "June Kim (jkim654)"
date: "4/15/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Exploring dataset

# Season Analysis 
## What is the percentage of wins in each game location - Home (H), Away (A) or Neutral (N)?
```{r}
  RegularSeasonDetailedResults = read.csv("PrelimData2018/RegularSeasonDetailedResults_Prelim2018.csv")
  Teams = read.csv("DataFiles/Teams.csv")
  library(plyr)
  library(ggplot2)

  # calculate frequency
  loc_freq = count(RegularSeasonDetailedResults$WLoc)
  
  # calculate percentage and round to 2sf
  loc_freq['percentage'] = 100/sum(loc_freq$freq)*loc_freq$freq
  loc_freq['percentage_r'] <- signif(loc_freq$percentage, digits = 3)
  
  # calculate center of each segment for label
  loc_freq['pos'] <- cumsum(loc_freq$percentage) - loc_freq$percentage/2

  # set levels for segments so can order to match labels
  loc_freq$x <- factor(loc_freq$x, levels = rev(loc_freq$x))
  
  # pie chart
  ggplot(loc_freq, aes(x="", y=percentage_r, fill = x))+
    geom_bar(stat="identity")+coord_polar("y")+
    ggtitle("% of games won at each location")+
    geom_text(aes(y=pos, label=percentage_r))+theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      axis.ticks = element_blank(),
      axis.text = element_blank(),
      panel.border = element_blank(),
      panel.grid = element_blank()
    )
```
We can definitely observe that nearly 60% of the wins occur at the winning teams' home stadium. 

# Win percentage for all teams
Are the same winners always winning? What were the top/least winning teams across all seasons? March Madness involves 68 Teams in total, so let's see who are top/worst 34 teams. 

```{r}
# win freq
team_Wfreq = count(RegularSeasonDetailedResults$WTeamID)
team_Wfreq$x = as.factor(team_Wfreq$x)

# lose freq
team_Lfreq = count(RegularSeasonDetailedResults$LTeamID)
team_Lfreq$x  = as.factor(team_Lfreq$x)

# rename common name to merge
colnames(team_Wfreq)[1] = "TeamID"
colnames(team_Lfreq)[1] = "TeamID"
team_Wfreq = merge(team_Wfreq, Teams, by = "TeamID")
win_freq = merge(team_Wfreq,team_Lfreq, by = "TeamID")
final_frequency = win_freq[-4:-5]
colnames(final_frequency)[2] = "Win"
colnames(final_frequency)[4] = "Lose"

# calculate win percentage
final_frequency["percentage"] = (100 / (final_frequency$Win + final_frequency$Lose)) * final_frequency$Win
final_frequency$TeamID <- as.factor(final_frequency$TeamID)

final_frequency$TeamID <- factor(final_frequency$TeamID, levels = final_frequency$TeamID[order(final_frequency$percentage)])

## Question: What are top 34 teams? 
top.teams = head(final_frequency[order(-final_frequency$percentage),], n=34)

## Question: What are worst 34 teams?
bottom.teams = head(final_frequency[order(final_frequency$percentage),], n=34)

# plot
 ggplot(top.teams, aes(x=TeamName, y=percentage)) +
  ylim(0, 100) +
  ggtitle("Top 34 teams % of games won") +
  geom_bar(stat='identity') +
  xlab("Team Name") + theme(
    axis.text.x = element_text(angle=90, hjust=1)
  )
```
Again, it is most likely that these 34 teams would be the winners in the future seasons as well. 

# What were some of the best/worst performing team overall throughout the seasons? 
```{r}
Wscores = as.data.frame(RegularSeasonDetailedResults$WTeamID)
colnames(Wscores)[1] = "TeamID"
Wscores["Score"] = RegularSeasonDetailedResults$WScore

Lscores = as.data.frame(RegularSeasonDetailedResults$LTeamID)
colnames(Lscores)[1] = "TeamID"
Lscores["Score"] = RegularSeasonDetailedResults$LScore

AllScores = cbind(Wscores,Lscores)

# calculate team with highest score
AggScores = aggregate(Score ~ TeamID, AllScores,FUN="mean")
AggScores = merge(AggScores, Teams, by = "TeamID")
AggScores$TeamID = as.factor(AggScores$TeamID)
AggScores$TeamID = factor(AggScores$TeamID, levels = AggScores$TeamID[order(AggScores$Score)])
AggScores = AggScores[,-4:-5]

# Top 34 scoring teams
head(AggScores[order(-AggScores$Score), ], n=34)

# Worst 34 scoring teams
head(AggScores[order(AggScores$Score), ], n=34)
```

## Question: Does higher average point score mean more winning? That is, is points scored correlated to the game result?
```{r}
Wscor <- as.data.frame(RegularSeasonDetailedResults$WTeamID)
colnames(Wscor)[1] <- "TeamID"
Wscor['Score'] <- RegularSeasonDetailedResults$WScore
Wscor['Result'] <- 'Win'

Lscor <- as.data.frame(RegularSeasonDetailedResults$LTeamID)
colnames(Lscor)[1] <- "TeamID"
Lscor['Score'] <- RegularSeasonDetailedResults$LScore
Lscor['Result'] <- 'Lose'

AllScores2 = rbind(Lscor,Wscor)

# Game result relative to total scores
ggplot(AllScores2, aes(x=Score,y=Result,fill=Result,colour=Result)) +
  xlim(0,150) + ggtitle("Result of the match to points scored") +
  geom_point(stat="identity")+
  xlab("Point Scored in Game")+
  ylab("Result")

# Calculate win probability based on scores
Wscores_freq <- as.data.frame(count(Wscor$Score))
Lscores_freq <- as.data.frame(count(Lscor$Score))
Scores_prob <- merge(Wscores_freq, Lscores_freq, by=c("x"="x"), all=TRUE)
colnames(Scores_prob)<- c("score","win","lose")

# function for calculating probability
probability <- function(x){
     x$win / (x$win + x$lose)
}
Scores_prob['probability'] <- probability(Scores_prob)

# What is the relationship in graph?
plot(Scores_prob$score,Scores_prob$probability,pch=19,xlab="Score",ylab="Probability of winning",main="Chances of winning relationship with Scores",col="darkblue")
```
Some of the matches had a situation where they scored high but ended up losing. 

# Win probability across features
How influential are each features on the result of the game?

```{r,warning=FALSE,message=FALSE}

library(gridExtra)

# pts = points scored in game
# fgm = 
# fgm3 = 
# fga = 
# fga3 = 
# ftm = 
# fta = 
# ast = assist
# or = 
# to = 
# stl = steal
# blk = block
# pf = 

# win probability based on fgm
wgfm = as.data.frame(count(RegularSeasonDetailedResults$WFGM))
lgfm = as.data.frame(count(RegularSeasonDetailedResults$LFGM))
fgm_prob = merge(wgfm,lgfm,by=c("x"="x"),all=T)
names(fgm_prob) = c("fgm","win","lose")
fgm_prob["prob"] = probability(fgm_prob)

# win probability based on fgm3
wgfm3 = as.data.frame(count(RegularSeasonDetailedResults$WFGM3))
lgfm3 = as.data.frame(count(RegularSeasonDetailedResults$LFGM3))
fgm3_prob = merge(wgfm3,lgfm3,by=c("x"="x"),all=T)
names(fgm3_prob) = c("fgm3","win","lose")
fgm3_prob["prob"] = probability(fgm3_prob)

# win probability based on fga 
wfga = as.data.frame(count(RegularSeasonDetailedResults$WFGA))
lfga = as.data.frame(count(RegularSeasonDetailedResults$LFGA))
fga_prob = merge(wfga,lfga,by=c("x"="x"),all=T)
names(fga_prob) = c("fga","win","lose")
fga_prob["prob"] = probability(fga_prob)

# win probability based on fga3
wfga3 = as.data.frame(count(RegularSeasonDetailedResults$WFGA3))
lfga3 = as.data.frame(count(RegularSeasonDetailedResults$LFGA3))
fga3_prob = merge(wfga,lfga,by=c("x"="x"),all=T)
names(fga3_prob) = c("fga3","win","lose")
fga3_prob["prob"] = probability(fga3_prob)

# win probability based on ftm
wftm = as.data.frame(count(RegularSeasonDetailedResults$WFTM))
lftm = as.data.frame(count(RegularSeasonDetailedResults$LFTM))
ftm_prob = merge(wftm,lftm,by=c("x"="x"),all=T)
names(ftm_prob) = c("ftm","win","lose")
ftm_prob["prob"] = probability(ftm_prob)

# win probability based on fta
wfta = as.data.frame(count(RegularSeasonDetailedResults$WFTA))
lfta = as.data.frame(count(RegularSeasonDetailedResults$LFTA))
fta_prob = merge(wfta,lfta,by=c("x"="x"),all=T)
names(fta_prob) = c("fta","win","lose")
fta_prob["prob"] = probability(fta_prob)

# win probability based on ast
wast = as.data.frame(count(RegularSeasonDetailedResults$WAst))
last = as.data.frame(count(RegularSeasonDetailedResults$LAst))
ast_prob = merge(wast,last,by=c("x"="x"),all=T)
names(ast_prob) = c("ast","win","lose")
ast_prob["prob"] = probability(ast_prob)

# win probability based on or
wor = as.data.frame(count(RegularSeasonDetailedResults$WOR))
lor  = as.data.frame(count(RegularSeasonDetailedResults$LOR))
or_prob = merge(wor,lor,by=c("x"="x"),all=T)
names(or_prob) = c("or","win","lose")
or_prob["prob"] = probability(or_prob)

# win probability based on to
wdr = as.data.frame(count(RegularSeasonDetailedResults$WDR))
ldr = as.data.frame(count(RegularSeasonDetailedResults$LDR))
dr_prob = merge(wdr,ldr,by=c("x"="x"),all=T)
names(dr_prob) = c("dr","win","lose")
dr_prob["prob"] = probability(dr_prob)

# win probability based on to
wto = as.data.frame(count(RegularSeasonDetailedResults$WTO))
lto = as.data.frame(count(RegularSeasonDetailedResults$LTO))
to_prob = merge(wto,lto,by=c("x"="x"),all=T)
names(to_prob) = c("to","win","lose")
to_prob["prob"] = probability(to_prob)

# win probability based on stl
wstl = as.data.frame(count(RegularSeasonDetailedResults$WStl))
lstl = as.data.frame(count(RegularSeasonDetailedResults$LStl))
stl_prob = merge(wstl,lstl,by=c("x"="x"),all=T)
names(stl_prob) = c("stl","win","lose")
stl_prob["prob"] = probability(stl_prob)

# win probability based on blk
wblk = as.data.frame(count(RegularSeasonDetailedResults$WBlk))
lblk = as.data.frame(count(RegularSeasonDetailedResults$LBlk))
blk_prob = merge(wblk,lblk,by=c("x"="x"),all=T)
names(blk_prob) = c("blk","win","lose")
blk_prob["prob"] = probability(blk_prob)

# win probability based on pf
wpf = as.data.frame(count(RegularSeasonDetailedResults$WPF))
lpf = as.data.frame(count(RegularSeasonDetailedResults$LPF))
pf_prob = merge(wpf,lpf,by=c("x"="x"),all=T)
names(pf_prob) = c("pf","win","lose")
pf_prob["prob"] = probability(pf_prob)

#---graph---#

p1 <- ggplot(Scores_prob, aes(x=score, y=probability)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("Points Scored in Game") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p2 <- ggplot(fgm_prob, aes(x=fgm, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("fgm") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p3 <- ggplot(fga_prob, aes(x=fga, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("fga") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p4 <- ggplot(fgm3_prob, aes(x=fgm3, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("fgm3") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p5 <- ggplot(fga3_prob, aes(x=fga3, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("fga3") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p6 <- ggplot(ftm_prob, aes(x=ftm, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("ftm") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p7 <- ggplot(fta_prob, aes(x=fta, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("fta") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p8 <- ggplot(or_prob, aes(x=or, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("or") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p9 <- ggplot(dr_prob, aes(x=dr, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("dr") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p10 <- ggplot(ast_prob, aes(x=ast, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("ast") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p11 <- ggplot(to_prob, aes(x=to, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("to") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p12 <- ggplot(stl_prob, aes(x=stl, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("stl") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p13 <- ggplot(blk_prob, aes(x=blk, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("blk") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

p14 <- ggplot(pf_prob, aes(x=pf, y=prob)) +
  geom_point(stat='identity') +
  geom_smooth() +
  ylim(0, 1) +
  xlab("pf") + 
  ylab('Win Probability') +
  theme(axis.title.x = element_text(size=11), axis.title.y = element_text(size=11)) +
  theme(aspect.ratio=1)

grid.arrange(p1, p2, p3, p4, p5,p6)
grid.arrange(p7,p8,p9,p10,p11,p12)
grid.arrange(p13,p14)
```

> Modeling

I am going to try using variables: points, fgm, ftm, fgm3, fta, dr, ast, stl, blk for logistic regression as they seem to be strong indicators of winning. 
